---
title: ""
author: ""
date: ""
output: 
  html_document:
    code_folding: show
    code_download: true
    toc: true                  # table of content true
    toc_depth: 3               # upto three depths of headings (specified by #, ## and ###)
    toc_float: true
    number_sections: true     # if you want number sections at each table header
    theme: united              # many options for theme, this one is my favorite.
    highlight: tango           # specifies the syntax highlighting style
    #css: my.css               # you can add your custom css, should be in same folder
---

# Set up misc {.tabset}
 
## remove scientific notation in printing 
```{r}
options(scipen=999)
```


## Download dependencies

```
# InstaLoad packages (update none of the dependencies)

# Download

## Bioconductory packages
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
  BiocManager::install("Biobase")

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
  BiocManager::install("CLL")

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("limma")

## util packages (EDA, dataviz)
if (!require("skimr")) install.packages("skim")
if (!require("devtools")) install.packages("devtools")
if (!require("dplyr")) install.packages("dplyr") 
if (!require("RColorBrewer")) install.packages("RColorBrewer") 
if (!require("ggplot2")) install.packages("ggplot2") 
if (!require("ggrepel")) install.packages("ggrepel") 
if (!require("ggiraph")) install.packages("ggiraph") 
if (!require("plotly")) install.packages("plotly") 

```


## Load packages
```{r}
# Load Bioconductor 
library(Biobase)    # for ExpressionSet object tidyness
library(CLL)        # for our example dataset, data(sCLLex)
library(limma)      # for differential expression analysis 
  
# Load Util Packages
library(skimr)
library(devtools)
library(dplyr)
library(RColorBrewer)
library(ggplot2)
library(ggrepel)
library(ggiraph)
library(plotly)

```


# Bioconductor `ExrpessionSet` from [{CLL}](https://bioconductor.org/packages/release/data/experiment/html/CLL.html)

```{r}
data("sCLLex")
eset<- sCLLex
dim(exprs(sCLLex))
# [1] 12625    22
summary(pData(sCLLex))
# SampleID             Disease  
# Length:22          progres.:14  
# Class :AsIs        stable  : 8  
# Mode  :character 
```

# Creating design matrix for `limma` model 

The outcome or y or predictor variable is the `Disease` status `{stable, progressing}` (binary outcome, aka control case)

```{r}
# Design matrix
design <- model.matrix(~Disease  , data = pData(eset)) 
summary(design)
 #  (Intercept) Diseasestable   
 # Min.   :1    Min.   :0.0000  
 # 1st Qu.:1    1st Qu.:0.0000  
 # Median :1    Median :0.0000  
 # Mean   :1    Mean   :0.3636  
 # 3rd Qu.:1    3rd Qu.:1.0000  
 # Max.   :1    Max.   :1.0000  
```

## Sanity checks for design matrix `dim()` and context
 
```{r}
head(design, 2) 
colSums(design) 
skimr::skim(pData(eset))
```


```{r}
# 2. Fit the coef of the model with lmfit by passing it the ExressionSet object and design matrix
fit <- limma::lmFit(eset, design)

# # 3. Calculate the t-statistics by passing the fitted model object to eth eBayes function
eBayes_fit <- limma::eBayes(fit)
class(eBayes_fit)
```



# Overview of significantly down and upregulated features with `decideTests` and `summary()`

```{r}
# the defaults from limma::decideTests() but explicit always better than implicit
FDR_threshold   <- 0.05
logFC_threshold <- 0
  
# Call the function limma::decideTests() with the `eBayes_fit` object as input
Down_NotSig_Up  <- limma::decideTests(
                                      object        = eBayes_fit,     # class(eBayes_fit): MArrayLM 
                                      method        = "separate",     # one coeff anyway, "Diseasestable", but check ??
                                      adjust.method = "BH"          , # Benjamini-Hochberg, adjusting for FDR
                                      p.value       = FDR_threshold , # family-wise error rate or FDR threshold
                                      lfc           = 0               # leave as is, default. Else inflation of tests
                                      )  
```

## `summary(limma::decideTests(ebayes_fit))`: Counts of down, up or not signif features absed on FDR threshold

```{r}
cat("Overview of statistically significant features for selected FDR (",FDR_threshold,"), on examined contrast ('stable','progress'):\n\n")
summary(Down_NotSig_Up)

# and some helper variable to retrieve this in a accessible dataframe:
summary_signif <- as.data.frame.matrix(t(summary(Down_NotSig_Up)))

cat("\n\nNumber of statistically significant features in total (regardless of direction of change): \n")
# > rownames(summary_signif)" [1] "(Intercept)"   "Diseasestable"
nSignif         <- summary_signif["Diseasestable","Down"] + summary_signif["Diseasestable", "Up"]
nTotalFeatures  <- sum(summary_signif["Diseasestable",])
cat(nSignif)
```


With FDR set to `r FDR_threshold` (soft-coded to inherit `FDR_threshold` value set above), `r `  differentially regulated features found between patients with `(Disease)stable` and `progres` subject of teh experiment for the examined contrast ()


# Table of ~~Top~~ all features from Linear Model Fit  (N = `r nTotalFeatures` features) from `topTable` object

For topTable, `fit`` should be an object of class `MArrayLM`
as produced by `lmFit` and `eBayes`.


```{r}
topTable_all <- limma::topTable(fit           = eBayes_fit,           #
                                coef          = "Diseasestable"    ,  # 
                                number        = dim(exprs(eset))[1],  # maximum features to list. List all
                                adjust.method = "BH"               ,  # adjust the p-values for multiple testing
                                p.value       = 1                  ,  # cutoff value for adjusted p-values. Show all
                                lfc           = 0)                    # cutoff value for log-fold change. Show all
```


## Check for NA in `P.values` and `adj.Pvalues` and replace those with `1`
```{r}
#replacing NA p-values
topTable_all$P.Value[is.na(topTable_all$P.Value)]   <- 1; 
topTable_all$adj.P.Val[is.na(topTable_all$adj.P.Val)] <- 1;
```

## Verify `decideResults()` output matches `topTable()` 

`sum(topTable_all$adj.P.Val < FDR_threshold)`: 
Count how many fetaures have adjusted bH p-value < `r FDR_threshold`

```{r}
sum(topTable_all$adj.P.Val < FDR_threshold)
```


# Volcano plot prep time - Creating helper variables for in-control `ggplot2` plotting 

HOW:
```
0. Use a custom palette
1. Use the `decideTests()` table 
 - Store output of `decideTests` in dataframe 
 - Create helper column `probe`, with same column name as in `toPlot` dataframe (more later on)
 - Remove `(Intercept)` column by setting to `NULL`
 - Use the `Diseasestable` column, but first, `as.factor` (ggplot2 loves these thingies)!

2. Map `Diseasestable` factor levels to specific hexcode using `match()`

3. Perform a `dplyr::inner_join()`
```

## Custom palette

```{r}
devtools::source_url("https://raw.githubusercontent.com/cgpu/ggmisc/master/web-palette.R")
webPalette()
```


## Create factor with 3 levels (`Down`, `Up`, `NotSignificant`)


- Use the `decideTests()` table 
   + Store output of `decideTests` in dataframe 
   + Create helper column `probe`, with same column name as in `toPlot` dataframe (more later on)
   + Remove `(Intercept)` column by setting to `NULL`
   + Use the `Diseasestable` column, but first, `as.factor` (ggplot2 loves these thingies)!
  
```{r}
# 1. Store output of `decideTests` in dataframe
Down_NotSig_Up_dataframe               <- as.data.frame.matrix(Down_NotSig_Up)
Down_NotSig_Up_dataframe$probe         <-  rownames(Down_NotSig_Up_dataframe)
Down_NotSig_Up_dataframe$`(Intercept)` <- NULL
Down_NotSig_Up_dataframe$Diseasestable <- as.factor(Down_NotSig_Up_dataframe$Diseasestable)

levels(Down_NotSig_Up_dataframe$Diseasestable )
```


## Map `Diseasestable` factor levels to specific hexcode from custom palette 



HOW: using `match()`

```{r}
my_ggpalette <- c(webPalette("watermelon fail"),
                  webPalette("cta blue"),
                  webPalette("grey dark"))

# gem found here at SO: 
# https://stackoverflow.com/questions/7658316/create-new-column-based-on-4-values-in-another-column
index  <- levels(Down_NotSig_Up_dataframe$Diseasestable )
values <- my_ggpalette

Down_NotSig_Up_dataframe$hexcode <- values[match(Down_NotSig_Up_dataframe$Diseasestable, index)]
head(Down_NotSig_Up_dataframe)
```


- Inspect dataframe and check what variables are missing

```{r}
toPlot <- topTable_all
head(topTable_all, 2)
```


## Create helper variable for visualizng small p-values in the positive side of the `y` axis.
```
  negative logPvalues `-1* (log10(FDR_threshold))` == `r -1* (log10(FDR_threshold))`
```

```{r}
# Negative log10 adjusted p-values
FDR_threshold_neg_log10 <-  (-1 * (log10(FDR_threshold)))
toPlot$neg_logPvalues <- -1* (log10(toPlot$adj.P.Val))

cat("Checking by condition:\n")
cat("sum(toPlot$neg_logPvalues > FDR_threshold_neg_log10 ) :\n")
cat(sum(toPlot$neg_logPvalues > FDR_threshold_neg_log10 ))
```

## Create a boolean vector column, significant TRUE/FALSE (binary)
```{r}
# Significant boolean vector
toPlot$significant <- toPlot$neg_logPvalues >= FDR_threshold_neg_log10


cat("\n\nVerifying helper variable:\n")
cat("sum(toPlot$significant == TRUE) :\n")
cat(sum(toPlot$significant == TRUE))
```

## Create variable that will facilitate `dplyr::inner_join()` with the `decideTests()` output dataframe
```{r}
# Assign rownames (probe ID) to a column
toPlot$probe <- rownames(toPlot)

cat("\ncolnames(toPlot)\n\n")
colnames(toPlot)
```

Inspect dataframe for plotting 

```{r}
head(toPlot, 4)
```

```{r}
skimr::skim(toPlot)
```


# Merge helper variable columns from `decideTests` output to `toPlot` dataframe with `merge()`

To use `merge()` make sure the dataframes:

- have the same rownames
- (and/or) share the same column

```{r}
cat("Is it TRUE that the two dataframes have exactly the same rownames (order is not signif):\n")
cat(setequal(rownames(Down_NotSig_Up_dataframe),rownames(toPlot)))
```

```{r}
colnames(Down_NotSig_Up_dataframe)
```

```{r}
colnames(toPlot)
```


```{r}
augmented_toPlot <- merge(Down_NotSig_Up_dataframe, toPlot)

dim(augmented_toPlot)
head(augmented_toPlot)
```
```{r}
colnames(augmented_toPlot)
```

```{r}
lower_custom_legend_label      = "lower"
not_signif_custom_legend_label = "no diff"
higher_custom_legend_label     = "higher"


minimal_toPlot <- augmented_toPlot[,c("probe", "hexcode", "logFC", "neg_logPvalues", "significant", "Diseasestable")]
minimal_toPlot$probe         <- as.factor(minimal_toPlot$probe)
minimal_toPlot$probe         <- as.factor(minimal_toPlot$probe)
minimal_toPlot$expression_dummy_variable   <- as.numeric(as.character(minimal_toPlot$Diseasestable))
minimal_toPlot$Diseasestable <- NULL


toPlot <- minimal_toPlot[order(minimal_toPlot$expression_dummy_variable),]
toPlot$expression_dummy_variable <- as.factor(toPlot$expression_dummy_variable)


# P0 this is unpredicatble must refactor to map & match color to value of expression column
index  <- levels(toPlot$expression_dummy_variable )
values <- c(lower_custom_legend_label, 
            not_signif_custom_legend_label,
            higher_custom_legend_label)

toPlot$expression <- values[match(toPlot$expression_dummy_variable, index)]
toPlot$expression <- as.factor(toPlot$expression)
toPlot$tooltip    <- toPlot$probe
# toPlot$expression_dummy_variable <- NULL

head(toPlot)
```
# Convert `toPlot` dataframe to  tible using `tibble::tibble(toPlot)`

```{r}
library(tibble)
df_ToPlot <- toPlot
toPlot <- tibble::as_tibble(df_ToPlot)
head(toPlot)
class(toPlot)
```


# Time for `ggplot2` 

```{r}
# Cutoff values for dashed lines
neg_logPvalue_threshold   <-  ((-1 * log10(FDR_threshold)))  
logFC_signif_cutoff       <- 1.5

# Creating helper variables for an airy plot, wide and centered plot
x_absolut_extremum        <- max(abs(toPlot$logFC))
x_absolut_extremum_wider  <- x_absolut_extremum * 1.65

y_absolut_extremum        <- max(abs(toPlot$neg_logPvalues))
y_absolut_extremum_wider  <- y_absolut_extremum * 1.15

ggtitle_custom            <- paste0("Volcano plot for genes in {CLL} ExpressionSet")
ggsubtitle_custom         <- paste0( nSignif, " features of ", nTotalFeatures, " found statistically significant, ( FDR =  ", FDR_threshold, ", BH adjustment)")
ggcaption_custom          <- paste0("logFoldChange signif cutoff: "  , logFC_signif_cutoff, "\n", 
                                    "- log AdjPvalue signif cutoff: ", round(neg_logPvalue_threshold, 2))

# Custom colours
ggtitle_custom_color       <- "#4A637B" # webPalette("grey dark")
ggsubtitle_custom_color    <- "#A9BACA" # webPalette("grey medium")
dashed_line_color          <- "#990000" # wine red
dashed_line_colour         <- "#990000" # wine red
scale_color_manual_palette <- c("#0198F5","#FA5E5B",  "#4A637B") 


x_axis_ticks_count = 10
y_axis_ticks_count = 10
  



# Start building gglayers of plotness
p<-ggplot() +   theme_gray()+
  geom_point(data    = toPlot, 
             mapping = aes(x       = logFC, 
                           y       = neg_logPvalues,
                           shape   = significant,
                           size    = significant,
                           color   = expression,
                           label   = probe,
                           tooltip = tooltip),
             alpha = 0.70)   +
  
  # Control legend block that shows first, here we choose 1, aka top position
  guides(color = guide_legend(order = 1)) +
  
  scale_color_manual(values  = scale_color_manual_palette ) +

  # geom_text_repel(aes(label = probe)) +
     
  xlab('logFoldChanges') +
  ylab('-log10(adjPValue)')      +
  

  # Set explicitly x axis ticks:https://stackoverflow.com/questions/11335836/increase-number-of-axis-ticks
  scale_x_continuous(breaks = pretty(toPlot$logFC         , n = x_axis_ticks_count)) +
  
  # Set ylim based on absolut extremum for achieving a centered volcano plot  
   ylim(0, y_absolut_extremum_wider)    +

  # Setting annotation for title, subtitle, caption
  labs(title      = ggtitle_custom,
       subtitle   = ggsubtitle_custom,
       caption    = ggcaption_custom)   +
  
  # Controlling font, colour and size of labs. Also centering with hjust 
  theme(
  plot.title    = element_text(color = ggtitle_custom_color, size = 16, face = "bold", hjust = 0.5),
  plot.subtitle = element_text(color = ggsubtitle_custom_color, size = 12             , hjust = 0.5),
  plot.caption  = element_text(color = ggsubtitle_custom_color, face = "italic")) +
  

  geom_text_repel(data      = subset(toPlot, significant == TRUE ), 
                  mapping   = (aes(x     = logFC,
                                   y     = neg_logPvalues,
                                   label = probe))) +
  
  geom_hline(yintercept =  neg_logPvalue_threshold     ,  colour= dashed_line_colour, linetype="dashed") + 
  geom_vline(xintercept =        logFC_signif_cutoff   ,  colour= dashed_line_colour, linetype="dashed") + 
  geom_vline(xintercept =  (-1 * logFC_signif_cutoff ) ,  colour= dashed_line_colour, linetype="dashed") 
  
# Add shadowing for double-filtering criterion, both stat signif adj.Pvalue and stat signif UPregulated
p <- p +  annotate ("rect", 
                    xmin = logFC_signif_cutoff, 
                    xmax = Inf, 
                    ymin = neg_logPvalue_threshold, 
                    ymax = Inf, alpha = .2) 

# Add another shadowing for double-filtering criterion, both stat signif adj.Pvalue and stat signif DOWNregulated
p <- p +  annotate ("rect", 
                    xmin = -logFC_signif_cutoff, 
                    xmax = -Inf, 
                    ymin = neg_logPvalue_threshold, 
                    ymax = Inf, alpha = .2)

htmlwidgets::saveWidget(ggplotly(p), file = "plotly_volcano.html")
```

